---
title: "About the Data"
subtitle: "Data sources, ingestion, cleaning, and exploratory analysis"
toc: true
toc-depth: 3
format:
  html:
    code-fold: true
    code-tools: true
execute:
  warning: false
  message: false
  echo: true
---

```{r setup}
#| include: false
Sys.setenv(LANGUAGE = "en")
options(repos = c(CRAN = "https://cloud.r-project.org"))
pkgs <- c("httr","jsonlite","dplyr","tidyr","ggplot2","sf","tigris","leaflet",
          "broom","knitr","RColorBrewer","rmapshaper","scales")
for (p in pkgs) {
  if (!requireNamespace(p, quietly = TRUE))
    suppressMessages(install.packages(p, quiet = TRUE))
}
invisible(lapply(pkgs, function(p) suppressPackageStartupMessages(library(p, character.only = TRUE))))
options(tigris_use_cache = TRUE)
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

# 1. Overview

This page documents every step of the data pipeline that feeds the main [Diabetes Report](diabetes.qmd). All data are public, downloaded live via APIs, and fully reproducible.

**Pipeline:**

```
CDC PLACES API ─┐
                ├─► join on 5-digit county FIPS ─► analytic dataset (N = 3,143)
Census ACS API ─┘
                                                   │
                      tigris county polygons ──────┴─► spatial layer
```

# 2. Raw Data Sources

## 2.1 CDC PLACES — Outcome Variable

**What it is.** PLACES (Population Level Analysis and Community Estimates) is the CDC's flagship small-area estimation program, publishing model-based prevalence estimates for 40+ chronic conditions and health behaviors at the county, census-tract, ZIP-code, and place level. Estimates are derived from the Behavioral Risk Factor Surveillance System (BRFSS) using multilevel post-stratification.

**Measure used.** `DIABETES` — "Diagnosed diabetes among adults aged ≥18 years," age-adjusted prevalence (%).

**Why this dataset.** CDC publishes several PLACES county releases, each with slightly different coverage due to BRFSS sample-size constraints. I tested four county-level datasets on Socrata and found only one with **complete 50-state coverage for age-adjusted diabetes**:

| Dataset ID | PA | KY | FL | CT | Total N | Selected? |
|---|---|---|---|---|---|---|
| `swc5-untb` (2024) | ❌ 0 | ❌ 0 | ✅ 67 | ✅ 9 | 2,958 | — |
| `h3ej-a9ec` (2021) | ✅ 67 | ✅ 120 | ❌ 0 | ✅ 8 | 3,077 | — |
| `cwsq-ngmh` | — | — | — | — | — | — |
| **`duw2-7jbt` (2020)** | **✅ 67** | **✅ 120** | **✅ 67** | **✅ 8** | **3,144** | **✅** |

**Endpoint (Socrata).** `https://data.cdc.gov/resource/duw2-7jbt.json`

**Query.** We pull only rows where `measureid = 'DIABETES'` and `datavaluetypeid = 'AgeAdjPrv'` (age-adjusted prevalence, not crude).

```{r pull-places}
soql <- "SELECT statedesc, stateabbr, locationname, locationid, data_value WHERE measureid='DIABETES' AND datavaluetypeid='AgeAdjPrv' LIMIT 50000"
url  <- paste0("https://data.cdc.gov/resource/duw2-7jbt.json?$query=",
               URLencode(soql, reserved = TRUE))

places_raw <- fromJSON(url)
places <- places_raw |>
  transmute(state = statedesc, state_abbr = stateabbr,
            county = locationname, fips = locationid,
            diabetes = as.numeric(data_value)) |>
  filter(!is.na(diabetes))

cat("Rows pulled from PLACES:", nrow(places), "\n")
cat("States represented:", length(unique(places$state_abbr)), "\n")
```

## 2.2 U.S. Census ACS — Covariates

**What it is.** The American Community Survey (ACS) 5-year estimates provide county-level socioeconomic indicators pooled over five survey years. We use the 2021 5-year release (covering 2017–2021) because it still uses Connecticut's traditional 8-county FIPS codes, which match PLACES; the 2022 release switched CT to "Planning Regions" with incompatible FIPS.

**Variables selected** (from ACS Subject Tables):

| Variable | ACS code | What it measures |
|---|---|---|
| `poverty_pct` | S1701_C03_001 | % of population below federal poverty line |
| `uninsured_pct` | S2701_C05_001 | % of population without health insurance |
| `median_income` | S1901_C01_012 | Median household income (USD) |
| `hs_or_higher_pct` | S1501_C02_014 | % of adults 25+ with a HS diploma or higher |

**Endpoint.** `https://api.census.gov/data/2021/acs/acs5/subject`

```{r pull-acs}
base <- "https://api.census.gov/data/2021/acs/acs5/subject"
vars <- "S1701_C03_001E,S2701_C05_001E,S1901_C01_012E,S1501_C02_014E"
resp <- GET(base, query = list(get = paste0("NAME,", vars),
                                `for` = "county:*", `in` = "state:*"))
acs <- as.data.frame(fromJSON(rawToChar(resp$content))[-1, ],
                     stringsAsFactors = FALSE)
names(acs) <- c("NAME","poverty_pct","uninsured_pct","median_income",
                "hs_or_higher_pct","state_fips","county_fips")
acs$GEOID <- paste0(acs$state_fips, acs$county_fips)
for (v in c("poverty_pct","uninsured_pct","median_income","hs_or_higher_pct"))
  acs[[v]] <- as.numeric(acs[[v]])

# Census API uses -666666666 (and other large negatives) as "jam values" for
# missing / suppressed data. Convert to NA so they don't contaminate summaries.
for (v in c("poverty_pct","uninsured_pct","median_income","hs_or_higher_pct"))
  acs[[v]][acs[[v]] < 0] <- NA

cat("ACS counties returned:", nrow(acs), "\n")
```

## 2.3 TIGER/Line County Boundaries

We use the `tigris` R package to download simplified county polygons (cartographic-boundary files, 20m resolution) for the 2021 vintage, matching the ACS year.

```{r pull-tigris, cache=TRUE}
counties_sf <- counties(cb = TRUE, resolution = "20m", year = 2021) |>
  filter(!STATEFP %in% c("02","15","60","66","69","72","78")) |>   # drop AK/HI/territories for Moran's I
  st_transform(4326) |>
  ms_simplify(keep = 0.3, keep_shapes = TRUE)

cat("County polygons (continental US):", nrow(counties_sf), "\n")
```

# 3. Data Cleaning and Processing

Every cleaning step below is a deliberate, documented decision. The goal is a single tidy analytic dataset where each row is one U.S. county and each column is either the outcome (diabetes prevalence) or a structural covariate.

## 3.1 Type Conversion

The Socrata and Census APIs return JSON arrays in which every value — even numeric ones — arrives as a **character string**. Before any arithmetic, we coerce the outcome and covariates to `numeric`:

```{r types, eval=FALSE}
# PLACES outcome
places$diabetes <- as.numeric(places$data_value)

# ACS covariates (same for all four)
for (v in c("poverty_pct","uninsured_pct","median_income","hs_or_higher_pct"))
  acs[[v]] <- as.numeric(acs[[v]])
```

Without this step, `mean()` and `cor()` silently return errors, and `ggplot` treats the values as discrete categories.

## 3.2 Long-to-Wide Format Decisions

The **PLACES raw download arrives in long format** — one row per county × measure × estimate type. A single county can appear dozens of times (once for each of ~40 measures: diabetes, hypertension, obesity, etc.). We do two things at once:

1. **Filter** to the single row per county that corresponds to `measureid = 'DIABETES'` AND `datavaluetypeid = 'AgeAdjPrv'` (age-adjusted prevalence, not crude).
2. **Select and rename** only the fields we need, producing a compact **wide-by-county** table ready for joining.

```{r tidy-places, eval=FALSE}
places <- places_raw |>
  transmute(state = statedesc, state_abbr = stateabbr,
            county = locationname, fips = locationid,
            diabetes = as.numeric(data_value))
```

The **ACS raw download arrives in long format by default** (one row per `(county, variable)`), but the Census API's `get=` parameter lets us request multiple variables in a single call, which the server returns as a **wide** table (one row per county, one column per variable). We rename the cryptic codes (`S1701_C03_001E` → `poverty_pct`, etc.) immediately so the downstream code is self-documenting.

## 3.3 Missing Data and Census "Jam Values"

The Census Bureau uses **special sentinel negative values** — the so-called "jam values" — to encode several kinds of unreported data (suppressed for confidentiality, insufficient sample, non-applicable geography). The most common is **`-666666666`**, but several other large negatives exist.

These jam values are **not missing in R's sense**: they are valid numeric entries that would silently poison every mean, SD, and correlation if left in. For example, Texas's Loving County (population ~60) had its median income reported as `-666666666` in our raw ACS pull — a single such row is enough to inflate the apparent range of median income to "−\$666M to \$156k" and drive the correlation between income and every other covariate to near zero.

We convert any negative jam value to `NA` before computing any statistics:

```{r jam, eval=FALSE}
for (v in c("poverty_pct","uninsured_pct","median_income","hs_or_higher_pct"))
  acs[[v]][acs[[v]] < 0] <- NA
```

After this step, the four covariates have plausible ranges (e.g., median income: \$17k–\$157k; poverty: 1%–59%) and the correlation matrix in §4.3 looks as expected.

## 3.4 Geographic Filtering

Several filters exclude geographies that are either not U.S. counties or incompatible with the spatial model:

| Filter | Where applied | Why |
|---|---|---|
| Drop `state = "United States"` row | PLACES | It is the national aggregate, not a county |
| Drop `STATEFP %in% c("60","66","69","72","78")` | TIGER (map + Moran's I) | American Samoa, Guam, Northern Mariana Islands, Puerto Rico, U.S. Virgin Islands — not in ACS county universe |
| Additionally drop `STATEFP %in% c("02","15")` | Moran's I only | Alaska and Hawaii have no land adjacency to the contiguous 48 states, so the queen-contiguity weights matrix would treat them as isolated islands with no neighbors |

## 3.5 Merge on 5-digit County FIPS

Both PLACES and ACS use the same 5-digit county FIPS code (`SSCCC`, where `SS` is the state FIPS and `CCC` is the county FIPS). We join on this key with an `inner_join`, then drop any row missing any covariate:

```{r merge}
merged <- places |>
  inner_join(acs, by = c("fips" = "GEOID")) |>
  filter(!is.na(diabetes), !is.na(poverty_pct), !is.na(uninsured_pct),
         !is.na(median_income), !is.na(hs_or_higher_pct)) |>
  mutate(income_10k = median_income / 10000)
```

The complete-case filter is **pairwise on the analytic set**, not imputation-based, because the fraction of missing counties is tiny (see the loss accounting below) and the missingness is plausibly MCAR (mostly sparsely-populated counties whose estimates fall below Census's reporting threshold).

## 3.6 Row-Loss Accounting

```{r losses}
n_places <- nrow(places)
n_acs    <- nrow(acs)
n_merged <- nrow(merged)
data.frame(
  Step = c("PLACES raw (counties + US aggregate)",
           "ACS raw (all U.S. counties)",
           "Inner join on FIPS + complete-case filter",
           "Final analytic dataset"),
  N = c(n_places, n_acs, n_merged, n_merged)
) |> knitr::kable()
```

The single row of attrition from PLACES → merged is the national aggregate (`FIPS = "US"`), which has no corresponding ACS county GEOID. The larger gap between ACS raw (~3,220 counties) and merged (3,143) reflects counties that exist in the Census universe but are missing from PLACES — typically because their BRFSS sample sizes were too small for a reliable model-based estimate that release year.

## 3.7 Variable Transformations

Two transformations happen in this notebook; a third happens in the main report.

| Variable | Transformation | Rationale |
|---|---|---|
| `income_10k` | `median_income / 10000` | Reports regression coefficients in "per \$10k" units instead of "per \$1" — more interpretable |
| (all four covariates) | z-score standardization in the main report | `(x − mean) / sd`; makes the four β coefficients directly comparable across variables with different units (%, \$, %, %) |
| Polygon simplification | `rmapshaper::ms_simplify(keep = 0.3)` on TIGER shapefile | Reduces vertex count to 30% for faster interactive map rendering; visual difference is negligible at country scale |

No outcome-side transformation is applied: the age-adjusted diabetes prevalence is already the reportable quantity, and the distribution (right-skewed but not log-normal; see §4.1) is compatible with ordinary least squares assumptions without a log or square-root transform.

# 4. Exploratory Data Analysis

## 4.1 Outcome Distribution

```{r fig-dist, fig.width=8, fig.height=4}
ggplot(merged, aes(diabetes)) +
  geom_histogram(bins = 50, fill = "#CB4B4B", color = "white") +
  geom_vline(xintercept = mean(merged$diabetes), linetype = 2) +
  labs(x = "Age-adjusted diabetes prevalence (%)",
       y = "Number of counties",
       title = "Distribution of County Diabetes Prevalence",
       subtitle = sprintf("N = %d · Mean = %.1f%% · SD = %.2f · Range = %.1f–%.1f%%",
                          nrow(merged), mean(merged$diabetes), sd(merged$diabetes),
                          min(merged$diabetes), max(merged$diabetes))) +
  theme_minimal(base_size = 12)
```

**Interpretation.** The distribution is right-skewed with a long upper tail of high-prevalence counties. Most counties cluster between 8% and 13%, but a small group exceeds 18% — these are concentrated in the rural South (Mississippi Delta, Alabama Black Belt), Appalachia, and Native American tribal lands. The 5.6%–21.5% range (nearly a 4-fold spread) signals that geographic disparities are substantial and warrant covariate modeling. The right skew also motivates using ordinary least squares with robust inference rather than a heavy transformation, since the tail is structural (specific high-burden regions), not statistical noise.

## 4.2 Covariate Distributions

```{r tbl-cov-summary}
#| tbl-cap: "Summary statistics for the four ACS covariates (N = 3,143 counties)."
merged |>
  summarise(
    `Poverty rate (%)`           = list(c(N = sum(!is.na(poverty_pct)),
                                          Mean = mean(poverty_pct),
                                          SD = sd(poverty_pct),
                                          Min = min(poverty_pct),
                                          Median = median(poverty_pct),
                                          Max = max(poverty_pct))),
    `Uninsured rate (%)`         = list(c(N = sum(!is.na(uninsured_pct)),
                                          Mean = mean(uninsured_pct),
                                          SD = sd(uninsured_pct),
                                          Min = min(uninsured_pct),
                                          Median = median(uninsured_pct),
                                          Max = max(uninsured_pct))),
    `Median household income ($)` = list(c(N = sum(!is.na(median_income)),
                                          Mean = mean(median_income),
                                          SD = sd(median_income),
                                          Min = min(median_income),
                                          Median = median(median_income),
                                          Max = max(median_income))),
    `% HS diploma or higher`     = list(c(N = sum(!is.na(hs_or_higher_pct)),
                                          Mean = mean(hs_or_higher_pct),
                                          SD = sd(hs_or_higher_pct),
                                          Min = min(hs_or_higher_pct),
                                          Median = median(hs_or_higher_pct),
                                          Max = max(hs_or_higher_pct)))
  ) |>
  tidyr::pivot_longer(everything(), names_to = "Variable", values_to = "stats") |>
  tidyr::unnest_wider(stats) |>
  dplyr::mutate(
    N      = format(N, big.mark = ","),
    Mean   = ifelse(grepl("income", Variable),
                    format(round(Mean, 0), big.mark = ","),
                    sprintf("%.2f", Mean)),
    SD     = ifelse(grepl("income", Variable),
                    format(round(SD, 0), big.mark = ","),
                    sprintf("%.2f", SD)),
    Min    = ifelse(grepl("income", Variable),
                    format(round(Min, 0), big.mark = ","),
                    sprintf("%.2f", Min)),
    Median = ifelse(grepl("income", Variable),
                    format(round(Median, 0), big.mark = ","),
                    sprintf("%.2f", Median)),
    Max    = ifelse(grepl("income", Variable),
                    format(round(Max, 0), big.mark = ","),
                    sprintf("%.2f", Max))
  ) |>
  knitr::kable(align = c("l", "r", "r", "r", "r", "r", "r"))
```

```{r fig-cov, fig.width=9, fig.height=6}
merged |>
  select(poverty_pct, uninsured_pct, median_income, hs_or_higher_pct) |>
  pivot_longer(everything(), names_to = "var", values_to = "val") |>
  mutate(var = recode(var,
    poverty_pct      = "Poverty rate (%)",
    uninsured_pct    = "Uninsured rate (%)",
    median_income    = "Median household income ($)",
    hs_or_higher_pct = "% with HS diploma or higher")) |>
  ggplot(aes(val)) +
  geom_histogram(bins = 40, fill = "#4B7BCB", color = "white") +
  facet_wrap(~ var, scales = "free", ncol = 2) +
  scale_x_continuous(labels = label_comma()) +
  labs(x = NULL, y = "Number of counties",
       title = "ACS Covariate Distributions Across U.S. Counties") +
  theme_minimal(base_size = 11)
```

**Interpretation.** The four covariates show very different shapes: poverty and uninsured rate are both right-skewed (most counties low, a tail of disadvantaged counties), median income is roughly log-normal with a few wealthy outliers (metro suburbs), and education (% HS or higher) is sharply left-skewed — the vast majority of U.S. counties now have over 85% high-school completion, and variation mostly lies in the lower tail (67%–90%). This asymmetry in the education variable means the "bad" end of the distribution (low HS completion) carries most of the explanatory signal for diabetes disparities.

## 4.3 Correlation Among Covariates

```{r fig-corr, fig.width=6, fig.height=5}
X <- merged |> select(poverty_pct, uninsured_pct, income_10k, hs_or_higher_pct)
cor_df <- as.data.frame(as.table(round(cor(X), 2)))
names(cor_df) <- c("Var1","Var2","r")

ggplot(cor_df, aes(Var1, Var2, fill = r, label = r)) +
  geom_tile(color = "white") +
  geom_text(size = 4) +
  scale_fill_gradient2(low = "#2C7BB6", mid = "white", high = "#D7191C",
                       midpoint = 0, limits = c(-1, 1)) +
  labs(x = NULL, y = NULL, title = "Covariate Correlation Matrix") +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        panel.grid = element_blank())
```

**Interpretation.** All four covariates are interrelated, as expected — they measure overlapping aspects of "structural disadvantage." The strongest pairwise correlation is between poverty and median income (r ≈ −0.72), reflecting the intuitive fact that poorer counties have lower median incomes. Poverty is also moderately correlated with education (r = −0.58) and with uninsured rate (r = 0.39), and education is moderately correlated with uninsured rate (r = −0.53). Critically, no pairwise correlation exceeds 0.75, and every variance-inflation factor in the main report is below 2.5 — well below the conventional multicollinearity threshold of 5. This justifies entering all four covariates simultaneously into the regression without dimension reduction.

## 4.4 Outcome vs. Covariate Scatterplots

```{r fig-scatter, fig.width=9, fig.height=7}
merged |>
  select(diabetes, poverty_pct, uninsured_pct, median_income, hs_or_higher_pct) |>
  pivot_longer(-diabetes, names_to = "var", values_to = "val") |>
  mutate(var = recode(var,
    poverty_pct      = "Poverty rate (%)",
    uninsured_pct    = "Uninsured rate (%)",
    median_income    = "Median household income ($)",
    hs_or_higher_pct = "% with HS diploma or higher")) |>
  ggplot(aes(val, diabetes)) +
  geom_point(alpha = 0.25, color = "#555", size = 0.8) +
  geom_smooth(method = "lm", color = "#CB4B4B", linewidth = 0.8) +
  facet_wrap(~ var, scales = "free_x", ncol = 2) +
  scale_x_continuous(labels = label_comma()) +
  labs(x = NULL, y = "Diabetes prevalence (%)",
       title = "Bivariate Relationships: Diabetes vs. ACS Covariates") +
  theme_minimal(base_size = 11)
```

**Interpretation.** All four predictors show a clear, monotonic association with diabetes prevalence in the theoretically expected direction: positive slopes for poverty and uninsured rate (more disadvantage → more diabetes), negative slopes for median income and educational attainment (more resources → less diabetes). Visually, the **poverty rate and education panels show the tightest linear relationships** — the clouds are narrow and the slopes are steep. The uninsured-rate and median-income panels also slope in the expected direction but with more scatter, consistent with their smaller coefficients in the multivariate regression. These bivariate patterns justify a linear model as a first approximation; any remaining non-linearity is flagged for robustness checks in the main report.

## 4.5 State-Level Averages

```{r fig-state-bar, fig.width=8, fig.height=8}
merged |>
  group_by(state_abbr, state) |>
  summarise(diabetes = mean(diabetes), n = dplyr::n(), .groups = "drop") |>
  ggplot(aes(reorder(state_abbr, diabetes), diabetes, fill = diabetes)) +
  geom_col() +
  scale_fill_distiller(palette = "YlOrRd", direction = 1) +
  coord_flip() +
  labs(x = NULL, y = "Mean county diabetes prevalence (%)",
       title = "State Averages — County Diabetes Prevalence") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "none")
```

**Interpretation.** The state-average ordering mirrors the classic Southeast "Diabetes Belt" first named by the CDC in 2011: Mississippi, Louisiana, Alabama, Georgia, and South Carolina at the top; Vermont, Massachusetts, Wyoming, New Hampshire, and Connecticut at the bottom. The visible gap between the top and bottom states — roughly 13% vs 7%, nearly a two-fold difference in mean prevalence — is striking on its own, but the state-average view hides substantial within-state heterogeneity (e.g., Texas has both some of the highest-burden border counties and some very low-burden metro suburbs). This is exactly why the main report complements this bar chart with a full **county-level choropleth** (§3.4 of the report) — the state view is an overview; the county view is the evidence.

# 5. Reproducibility

All code, data URLs, and package versions needed to recreate every figure and table on this page are contained in the source `data.qmd`. Click the **`</> Code`** button in the page header to view all code blocks.

```{r sessioninfo}
sessionInfo()
```
