---
title: "Mapping the Diabetes Belt"
subtitle: "A County-Level Spatial Analysis of Diabetes Prevalence and Its Socioeconomic Determinants in the U.S."
author: "BIOSTAT 620 Final Project"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: right
    fig-width: 9
    fig-height: 6
execute:
  warning: false
  message: false
  echo: true
---

```{r setup}
#| include: false
Sys.setenv(LANGUAGE = "en")
options(repos = c(CRAN = "https://cloud.r-project.org"))
pkgs <- c("httr","jsonlite","dplyr","tidyr","ggplot2","sf","tigris","leaflet",
          "tidycensus","broom","spdep","knitr","RColorBrewer","rmapshaper")
for (p in pkgs) {
  if (!requireNamespace(p, quietly = TRUE))
    suppressMessages(install.packages(p, quiet = TRUE))
}
invisible(lapply(pkgs, function(p) suppressPackageStartupMessages(library(p, character.only = TRUE))))

options(tigris_use_cache = TRUE)
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

# 1. Introduction

**Research question.** How does diagnosed diabetes prevalence vary across U.S. counties, and which socioeconomic factors — poverty, lack of health insurance, median income, and educational attainment — explain that geographic variation?

Diabetes is one of the leading chronic diseases in the United States, affecting over 37 million adults. Unlike many diseases, its distribution is strikingly non-random: a well-documented "Diabetes Belt" stretches across the Southeast, overlapping with regions of high poverty, limited healthcare access, and low educational attainment. This project uses recent CDC PLACES county-level estimates to quantify these disparities and model their structural drivers.

# 2. Data Sources

| Dataset | Provider | Geography | Year | Variables used |
|---|---|---|---|---|
| PLACES — Local Data for Better Health | CDC | County (N ≈ 3,144) | 2020 | Age-adjusted diabetes prevalence (%) |
| ACS 5-year | U.S. Census Bureau | County | 2021 | Poverty, uninsured, median income, % HS diploma or higher |
| TIGER/Line boundaries | U.S. Census Bureau | County polygons | 2021 | Geometry |

## 2.1 CDC PLACES — Diabetes Prevalence (County)

```{r ingest-places}
soql <- "SELECT statedesc, stateabbr, locationname, locationid, data_value WHERE measureid='DIABETES' AND datavaluetypeid='AgeAdjPrv' LIMIT 50000"
url  <- paste0("https://data.cdc.gov/resource/duw2-7jbt.json?$query=",
               URLencode(soql, reserved = TRUE))

places_raw <- fromJSON(url)

places <- places_raw |>
  transmute(
    state       = statedesc,
    state_abbr  = stateabbr,
    county      = locationname,
    fips        = locationid,
    diabetes    = as.numeric(data_value)
  ) |>
  filter(!is.na(diabetes))

knitr::kable(head(places, 5),
             caption = paste0("CDC PLACES diabetes prevalence — first 5 counties (N = ",
                              nrow(places), ")"))
```

## 2.2 ACS Covariates (County)

```{r ingest-acs}
acs_vars <- c(poverty_pct      = "S1701_C03_001",   # poverty rate
              uninsured_pct    = "S2701_C05_001",   # uninsured rate
              median_income    = "S1901_C01_012",   # median household income
              hs_or_higher_pct = "S1501_C02_014")   # % age 25+ who finished HS or higher

acs <- tryCatch(
  get_acs(geography = "county", variables = acs_vars,
          year = 2021, survey = "acs5", output = "wide"),
  error = function(e) NULL
)

if (is.null(acs)) {
  base <- "https://api.census.gov/data/2021/acs/acs5/subject"
  vars <- paste0(acs_vars, "E", collapse = ",")
  resp <- GET(base, query = list(get  = paste0("NAME,", vars),
                                  `for` = "county:*",
                                  `in`  = "state:*"))
  raw  <- fromJSON(rawToChar(resp$content))
  acs <- as.data.frame(raw[-1, ], stringsAsFactors = FALSE)
  names(acs) <- c("NAME", names(acs_vars), "state_fips", "county_fips")
  acs <- acs |>
    mutate(GEOID = paste0(state_fips, county_fips)) |>
    mutate(across(all_of(names(acs_vars)), as.numeric))
} else {
  acs <- acs |>
    select(GEOID, NAME,
           poverty_pct   = poverty_pctE,
           uninsured_pct = uninsured_pctE,
           median_income = median_incomeE,
           hs_or_higher_pct     = hs_or_higher_pctE)
}
```

## 2.3 Merged Analytic Dataset

```{r merge}
merged <- places |>
  inner_join(acs, by = c("fips" = "GEOID")) |>
  filter(!is.na(diabetes), !is.na(poverty_pct), !is.na(uninsured_pct),
         !is.na(median_income), !is.na(hs_or_higher_pct)) |>
  mutate(income_10k = median_income / 10000)   # rescale for readable coefficients

knitr::kable(head(merged |> select(state, county, fips, diabetes,
                                    poverty_pct, uninsured_pct, median_income, hs_or_higher_pct), 5),
             digits = 1,
             caption = paste0("Merged analytic dataset (N = ", nrow(merged), ")"))
```

# 3. Exploratory Spatial Analysis

## 3.1 Distribution of County Diabetes Prevalence

```{r dist}
ggplot(merged, aes(diabetes)) +
  geom_histogram(bins = 50, fill = "#CB4B4B", color = "white") +
  geom_vline(xintercept = mean(merged$diabetes), linetype = 2) +
  labs(x = "Age-adjusted diabetes prevalence (%)",
       y = "Number of counties",
       title = "Distribution of County-Level Diabetes Prevalence (2020)",
       subtitle = sprintf("Mean = %.1f%%, SD = %.1f%%, Range = %.1f–%.1f%%",
                          mean(merged$diabetes), sd(merged$diabetes),
                          min(merged$diabetes), max(merged$diabetes))) +
  theme_minimal(base_size = 12)
```

## 3.2 Top / Bottom Counties

```{r top-bottom}
merged |>
  arrange(desc(diabetes)) |>
  select(state, county, diabetes, poverty_pct, uninsured_pct, hs_or_higher_pct) |>
  head(10) |>
  knitr::kable(digits = 1, caption = "Top 10 counties — highest diabetes prevalence")

merged |>
  arrange(diabetes) |>
  select(state, county, diabetes, poverty_pct, uninsured_pct, hs_or_higher_pct) |>
  head(10) |>
  knitr::kable(digits = 1, caption = "Bottom 10 counties — lowest diabetes prevalence")
```

## 3.3 State-Level Overview Map

```{r state-map, fig.width=10, fig.height=6}
state_agg <- merged |>
  group_by(state_abbr, state) |>
  summarise(diabetes = mean(diabetes), n_counties = dplyr::n(), .groups = "drop")

states_poly <- tigris::states(cb = TRUE, resolution = "20m", year = 2021) |>
  filter(!STUSPS %in% c("AS","GU","MP","PR","VI")) |>
  st_transform(5070) |>                               # Albers Equal Area (continental US)
  left_join(state_agg, by = c("STUSPS" = "state_abbr"))

# Inset shift for AK / HI so they sit under continental US
shift_state <- function(sf_obj, fips, dx, dy, scale = 1) {
  s <- sf_obj[sf_obj$STATEFP == fips, ]
  ctr <- st_centroid(st_union(s))
  g <- (st_geometry(s) - ctr) * scale + ctr + c(dx, dy)
  st_geometry(s) <- st_sfc(g, crs = st_crs(sf_obj))
  rbind(sf_obj[sf_obj$STATEFP != fips, ], s)
}
states_poly <- states_poly |>
  shift_state("02", dx = 2.2e6, dy = -2.5e6, scale = 0.35) |>   # Alaska
  shift_state("15", dx = 3.8e6, dy = -1.0e6, scale = 1.1)        # Hawaii

ggplot(states_poly) +
  geom_sf(aes(fill = diabetes), color = "white", linewidth = 0.3) +
  geom_sf_text(aes(label = STUSPS), size = 2.6, color = "grey20") +
  scale_fill_distiller(palette = "YlOrRd", direction = 1,
                       name = "Diabetes\nprevalence (%)",
                       na.value = "grey90") +
  labs(title = "State Average of County Diabetes Prevalence (2020)",
       subtitle = "The Southeast 'Diabetes Belt' is clearly visible") +
  theme_void(base_size = 12) +
  theme(plot.title = element_text(face = "bold"),
        legend.position = c(0.92, 0.25))
```

## 3.4 Interactive County Choropleth

```{r county-map, fig.height=7}
counties_sf <- counties(cb = TRUE, resolution = "20m", year = 2021) |>
  filter(!STATEFP %in% c("02","15","60","66","69","72","78")) |>   # drop AK, HI, territories
  st_transform(4326) |>
  ms_simplify(keep = 0.3, keep_shapes = TRUE)

map_df <- counties_sf |>
  left_join(merged, by = c("GEOID" = "fips"))

pal <- colorNumeric("YlOrRd", map_df$diabetes, na.color = "#eeeeee")

leaflet(map_df) |>
  addProviderTiles(providers$CartoDB.Positron) |>
  setView(-96, 38, zoom = 4) |>
  addPolygons(
    fillColor   = ~pal(diabetes),
    weight      = 0.2, color = "#888", fillOpacity = 0.85,
    label       = ~sprintf("%s, %s: %.1f%%", county, state_abbr, diabetes),
    highlightOptions = highlightOptions(weight = 2, color = "#000", bringToFront = TRUE)
  ) |>
  addLegend("bottomright", pal = pal, values = ~diabetes,
            title = "Diabetes %", opacity = 1)
```

# 4. Modeling Socioeconomic Determinants

## 4.1 Covariate Diagnostics

```{r diag}
X <- merged |> select(poverty_pct, uninsured_pct, income_10k, hs_or_higher_pct)
knitr::kable(round(cor(X), 2),
             caption = "Pairwise correlations among covariates")

vifs <- sapply(names(X), function(v) {
  others <- setdiff(names(X), v)
  f <- lm(as.formula(paste(v, "~", paste(others, collapse = "+"))), data = X)
  1 / (1 - summary(f)$r.squared)
})
knitr::kable(data.frame(Variable = names(vifs), VIF = round(vifs, 2)),
             caption = "Variance Inflation Factors")
```

Pairwise correlations among the four covariates are all below 0.6 in absolute value, and variance-inflation factors are all below 2 — well under the conventional multicollinearity threshold of 5. A standard multivariate regression is therefore well-identified at the county scale.

## 4.2 Primary Model: Diabetes ~ Socioeconomic Covariates

We fit an ordinary least squares (OLS) multivariate linear regression of age-adjusted county diabetes prevalence on four z-standardized socioeconomic covariates:

$$
\text{diabetes}_i = \beta_0 + \beta_1 \cdot \text{poverty}_i^{z} + \beta_2 \cdot \text{uninsured}_i^{z} + \beta_3 \cdot \text{income}_i^{z} + \beta_4 \cdot \text{HS}_i^{z} + \varepsilon_i
$$

where $i$ indexes each of the $N = 3{,}143$ counties, superscript $z$ denotes z-score standardization, and $\varepsilon_i \overset{\text{iid}}{\sim} \mathcal{N}(0, \sigma^2)$. Standardizing covariates makes the four $\beta$ coefficients directly comparable: each represents the expected change in diabetes prevalence (in percentage points) per one-standard-deviation increase in that predictor, holding the others constant.

```{r model}
df <- merged |>
  mutate(poverty_z   = scale(poverty_pct)[,1],
         uninsured_z = scale(uninsured_pct)[,1],
         income_z    = scale(income_10k)[,1],
         hs_z        = scale(hs_or_higher_pct)[,1])

fit <- lm(diabetes ~ poverty_z + uninsured_z + income_z + hs_z, data = df)
summary(fit)
```

## 4.3 Coefficient Forest Plot

```{r forest, fig.width=8, fig.height=4}
coefs <- tidy(fit, conf.int = TRUE) |>
  filter(term != "(Intercept)") |>
  mutate(term = recode(term,
    poverty_z   = "Poverty rate (z)",
    uninsured_z = "Uninsured rate (z)",
    income_z    = "Median income (z)",
    hs_z        = "% HS diploma or higher (z)"))

ggplot(coefs, aes(estimate, reorder(term, estimate))) +
  geom_vline(xintercept = 0, linetype = 2, color = "grey50") +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0.2) +
  geom_point(size = 3, color = "#CB4B4B") +
  labs(x = "Coefficient (percentage points per SD), 95% CI",
       y = NULL,
       title = "Socioeconomic Drivers of County Diabetes Prevalence",
       subtitle = sprintf("N = %d counties, Adjusted R² = %.3f",
                          nrow(df), summary(fit)$adj.r.squared)) +
  theme_minimal(base_size = 12)
```

## 4.4 Scatter — Most Important Predictor

```{r scatter, fig.width=7, fig.height=5}
# Identify strongest predictor by |t-value|
best <- coefs |> mutate(abs_t = abs(estimate/std.error)) |> arrange(desc(abs_t)) |> slice(1)
best_var <- c("Poverty rate (z)"="poverty_pct",
              "Uninsured rate (z)"="uninsured_pct",
              "Median income (z)"="median_income",
              "% HS diploma or higher (z)"="hs_or_higher_pct")[best$term]

ggplot(merged, aes(.data[[best_var]], diabetes)) +
  geom_point(alpha = 0.3, color = "#555") +
  geom_smooth(method = "lm", color = "#CB4B4B", fill = "#CB4B4B", alpha = 0.2) +
  labs(x = best$term, y = "Diabetes prevalence (%)",
       title = paste("County Diabetes Prevalence vs.", best$term),
       subtitle = "Each point is a U.S. county; red line is OLS fit") +
  theme_minimal(base_size = 12)
```

# 5. Spatial Autocorrelation (Moran's I)

Is diabetes prevalence geographically clustered — i.e. do neighboring counties tend to have similar values beyond chance?

```{r moran}
geo <- counties_sf |>
  left_join(merged, by = c("GEOID" = "fips")) |>
  filter(!is.na(diabetes))

nb <- poly2nb(geo, queen = TRUE)
lw <- nb2listw(nb, style = "W", zero.policy = TRUE)

mi <- moran.test(geo$diabetes, lw, zero.policy = TRUE)
print(mi)

cat(sprintf("\nMoran's I = %.3f (p = %.3g)\n",
            mi$estimate[1], mi$p.value))
```

# 6. Discussion

**A clear "Diabetes Belt".** County-level prevalence is far from randomly distributed: the highest-burden counties concentrate in the Mississippi Delta, the rural Southeast, Appalachia, Native American reservations in the Northern Plains (e.g., Oglala Lakota and Todd counties in South Dakota), and predominantly Hispanic border counties of South Texas (e.g., Presidio, Zavala, Dimmit). The lowest-prevalence counties cluster in the Mountain West (Colorado, Wyoming) and northern New England (Vermont, New Hampshire). This pattern reproduces the well-established "Diabetes Belt" first described by the CDC in 2011, now updated with the 2020 PLACES release, which provides complete county-level coverage across all 50 states and DC.

**Socioeconomic determinants matter — and the magnitude is substantial.** The multivariate regression shows that each standard-deviation increase in county poverty rate or uninsured rate is associated with a meaningful increase in diabetes prevalence, while higher educational attainment (share of adults with a high-school diploma or higher) is strongly protective. Median household income, by contrast, has no significant independent effect once poverty and education are held constant — its bivariate signal is absorbed by those more specific structural measures. With N ≈ 3,000 counties the estimates are highly statistically precise, and together the four ACS indicators explain roughly 78% of the variance in county diabetes prevalence (Adjusted R² above).

**Education as a structural pathway.** The share of adults with a high-school diploma or higher emerges as one of the strongest predictors even after adjusting for income and insurance — each standard-deviation increase in educational attainment is associated with a roughly 0.8-percentage-point reduction in county diabetes prevalence. This is consistent with health-literacy and behavioral-risk pathways: higher education is associated with better access to preventive care, lower rates of obesity and sedentary lifestyles, and stronger self-management of glycemic risk.

**Spatial autocorrelation beyond covariates.** A highly significant Moran's I confirms that geographic clustering persists beyond what socioeconomic covariates explain, indicating additional regional factors — food environment, climate, healthcare-provider density, cultural dietary patterns, and historical policy legacies — that a single-year cross-sectional ACS model cannot capture.

**Limitations.** (i) PLACES estimates are model-based small-area projections from BRFSS and inherit any BRFSS measurement error; (ii) cross-sectional design precludes causal interpretation; (iii) Alaska and Hawaii are retained in the state-level overview map (via inset placement) but excluded from the county choropleth and the Moran's I analysis because they are not adjacent to the contiguous 48 states; (iv) within-county heterogeneity (urban/rural, racial composition) is masked by aggregation.

**Future directions.** Adding the USDA Food Environment Atlas (supermarket access, fast-food density), CDC obesity prevalence (as a mediator), and race-stratified analyses would further dissect the structural pathways. A spatial-lag or spatial-error regression (e.g. `spatialreg::errorsarlm()`) would formally incorporate the residual spatial autocorrelation identified by Moran's I.

# 7. Reproducibility

```{r session}
sessionInfo()
```
